---
title: "STAT631 Mini Project Report"
author: "David Teng"
date: "`r Sys.Date()`"
output: pdf_document
format: pdf
editor: visual
---

### Data Description

This analysis uses the Right Heart Catheterization (RHC) dataset, which includes 5,735 critically ill adult patients from the SUPPORT study (Study to Understand Prognoses and Preferences for Outcomes and Risks of Treatments), conducted from 1989 to 1994 across five U.S. teaching hospitals. The dataset is publicly available through the Vanderbilt University Department of Biostatistics. For the purpose of this study, the outcome of interest is hospital length of stay, defined as the number of days from admission to discharge. The independent variables examined are sex (male or female), age group (categorized as \<50, 50–65, 65–80, and 80+), and history of cardiovascular disease (yes or no).

### Right Heart Catheterization (RHC) Dataset Overview

Source: Vanderbilt University Department of Biostatistics

Direct Download: <https://hbiostat.org/data/repo/rhc.csv>

Dataset Documentation: <https://search.r-project.org/CRAN/refmans/ATbounds/html/RHC.html>

Study Reference: Connors, A.F., et al. (1996). "The effectiveness of right heart catheterization in the initial care of critically ill patients." JAMA, 276(11), 889–897. DOI: 10.1001/jama.1996.03540110043030

### Research Question

This study examines whether sex, age group, and history of cardiovascular disease (cardiohx) individually or interactively influence hospital length of stay among critically ill patients.

### Load Libraries and Data

```{r}
library(ggplot2)
library(dplyr)
library(multcomp)
library(stats)
library(car)  

df <- read.csv("rhc.csv")
summary(df)
```

### Data Preparation

```{r}
# Create age group categories from continuous age variable
df$age_group <- cut(df$age,
                    breaks = c(0, 50, 65, 80, 100),
                    labels = c("<50", "50-65", "65-80", "80+"),
                    right = FALSE)

# Convert categorical variables to factors
df$sex <- as.factor(df$sex)
df$cardiohx <- as.factor(df$cardiohx)

# Ensure death is a factor and create numeric binary version (if not already)
df$death <- as.factor(df$death)
df$death_num <- ifelse(df$death == "Yes", 1, 0)

# Select variables and drop rows with any missing values in selected columns
df_clean <- na.omit(df[, c("death", "sex", "age", "age_group", "cardiohx",
                           "death_num", "race", "income", "adld3p",
                           "urin1", "dschdte", "sadmdte")])

# Ensure date columns are in Date or DateTime format
df_clean$dschdte <- as.Date(df_clean$dschdte)
df_clean$sadmdte <- as.Date(df_clean$sadmdte)

# Calculate length of stay in days
df_clean$length_of_stay <- as.numeric(df_clean$dschdte - df_clean$sadmdte)

summary(df_clean)
```

### Assumption Checking Before ANOVA

1\. **Independence**

Assumed by study design (random sampling or independent subjects).\
No formal test—assume valid if no clustering/repeated measures.

2\. **Normality of Residuals**

```{r}
# Fit the model:
anova_model <- aov(length_of_stay ~ sex * age_group * cardiohx, data = df_clean)
# Check residual normality:
resid_anova <- residuals(anova_model)
shapiro.test(resid_anova)
qqnorm(resid_anova)
qqline(resid_anova, col = "blue")

```

Since p \< 0.05 **Normality of Residuals** is violated → Consider **transformation (e.g., log, sqrt) or** Logistic Regression

3\. **Equal Variance**

```{r}
# Levene's Test
library(car)
leveneTest(resid_anova ~ interaction(sex, age_group, cardiohx), data = df_clean)

# Fit the ANOVA model
anova_model <- aov(length_of_stay ~ sex * age_group * cardiohx, data = df_clean)

# Extract residuals and fitted values
resid_anova <- residuals(anova_model)
fitted_anova <- fitted(anova_model)

# Plot: Residuals vs. Fitted Values
plot(fitted_anova, resid_anova,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs. Fitted Values",
     pch = 20,
     col = "steelblue")
abline(h = 0, col = "red", lwd = 2)


```

**Interpretation**:

Levene p \> 0.05 → Variances are **equal**

Plot: Look for random scatter around 0 (no funnel pattern)

### Option 1: **Log Transformation**

```{r}
# Add a small constant to avoid log(0), if needed
df_clean$log_los <- log(df_clean$length_of_stay + 1)

# Fit the model using log-transformed outcome
anova_model_log <- aov(log_los ~ sex * age_group * cardiohx, data = df_clean)

# Check residuals again
resid_log <- residuals(anova_model_log)
shapiro.test(resid_log)
qqnorm(resid_log)
qqline(resid_log, col = "blue")

```

### Option 2: **Box-Cox Transformation**

```{r}
# Use lm instead of aov for Box-Cox compatibility
library(MASS)
lm_model <- lm(length_of_stay ~  sex * age_group * cardiohx, data = df_clean)

# Apply Box-Cox transformation
boxcox_result <- boxcox(lm_model, lambda = seq(-2, 2, 0.1),
                        main = "Box-Cox Transformation")

# Find optimal lambda
best_lambda <- boxcox_result$x[which.max(boxcox_result$y)]
best_lambda

```

Since `lambda ≈ 0`, then **log transformation** is best.

### check if the design is balanced 

```{r}
table(df_clean$sex, df_clean$age_group, df_clean$cardiohx)

```

Since this variability clearly indicates an **unbalanced design**, type II ANOVA is better suited.

### 1. Three-Way Type II ANOVA: Effects of Sex, Age Group, and Cardiac History on Hospital Length of Stay (Full Model)

```{r}
library(car)

# Log-transform length_of_stay (add 1 to avoid log(0))
df_clean$log_los <- log(df_clean$length_of_stay + 1)

# Fit the Type II ANOVA model (three-way design) using log-transformed length of stay
lm_full <- lm(log_los ~ sex * age_group * cardiohx, data = df_clean)
Anova(lm_full, type = 2)

```

### Key Findings (Full Model, Type II ANOVA):

The Type II ANOVA model, fitted using the log-transformed hospital length of stay, revealed that most main effects and two-way interactions were not statistically significant. Specifically, sex (p = 0.468), age group (p = 0.601), and cardiac history (p = 0.099) did not individually predict length of stay, nor did their respective two-way interactions involving sex.

The only statistically significant effect was the interaction between age group and cardiac history (F(3, 618) = 3.30, p = 0.020), indicating that the influence of cardiac history on hospital length of stay differs depending on the patient's age group. The three-way interaction among sex, age group, and cardiac history was not significant (p = 0.627), suggesting no combined effect across all three variables.

These results validate the choice to proceed with a reduced model focusing on age group and cardiac history, as sex and higher-order interactions did not contribute meaningfully to the explanation of variation in hospital length of stay.

### 2. Two-Way Type II ANOVA for Age Group × Cardiac History (Reduced Model)

Since the **three-way type II ANOVA** showed that only the **interaction between age group and cardiac history** is significant, it's appropriate to fit a **reduced Two-Way Type II ANOVA model** including just those two variables and their interaction.

```{r}
# Fit the Type II ANOVA model (two-way design) using log-transformed length of stay
lm_reduced <- lm(log_los ~ age_group * cardiohx, data = df_clean)
Anova(lm_reduced, type = 2)

```

### Key Findings (Reduced Model, Type II ANOVA):

In the reduced two-way ANOVA model using log-transformed hospital length of stay as the response variable, neither age group (p = 0.652) nor cardiac history (p = 0.094) demonstrated a significant main effect. However, the interaction between age group and cardiac history was statistically significant (F(3, 626) = 2.81, p = 0.0388).

This result indicates that the impact of cardiac history on hospital length of stay differs across age groups, or conversely, that the effect of age group on length of stay varies depending on whether or not the patient has a history of cardiovascular disease. The lack of main effects alongside a significant interaction suggests that the two variables do not independently influence length of stay but instead **interact** to shape outcomes in a joint and context-dependent manner.

### 
**ANOVA model comparison**

```{r}
anova(lm_reduced, lm_full)
```

This test if the reduced model provides a **significantly better fit** than the full odel. Since the p-value is **not significant** and Residual Sum of Squares (RSS) barely changed, then the **reduced model is sufficient**.

### AIC/ BIC Comparison

```{r}
# Compare AIC values
AIC(lm_reduced, lm_full)
# Compare BIC values
BIC(lm_reduced, lm_full)
```

### 

### Interpretation of AIC/ BIC comparisons:

The reduced model (which includes only age group, cardiac history, and their interaction) has a lower AIC than the full model by **11.57 points**. A difference in AIC greater than 10 is considered **strong evidence** in favor of the simpler model. Based on this AIC comparison, the reduced model is **strongly preferred** over the full model. Removing the non-significant terms involving sex improves **model parsimony** without sacrificing model performance.

Similarly, the **Bayesian Information Criterion (BIC)** further supports this conclusion. The reduced model has a BIC of **1281.28**, while the full model's BIC is **1328.47**, yielding a difference of over **47 points**. Since BIC penalizes model complexity more heavily than AIC, this substantial gap provides **very strong evidence** that the reduced model offers a more efficient and generalizable fit.

### Model Comparison Table 

##### (these are **linear models with a log-transformed response)**

| Model             | Residual DF | Residual SS | AIC     | BIC     | Nested ANOVA p-value |
|------------|------------|------------|------------|------------|------------|
| **Full Model**    | 618         | 253.80      | 1252.78 | 1328.47 | 0.8258               |
| **Reduced Model** | 626         | 255.58      | 1241.21 | 1281.28 | —                    |

### Model Decision:

The **reduced model** (including only age group, cardiac history, and their interaction) is preferred. It has **lower AIC and BIC values**, and the **ANOVA comparison** shows no significant loss of fit when the non-significant sex-related terms are removed (nested ANOVA p-value = 0.8258).

### Tukey HSD Post-Hoc Analysis of Age Groups on Hospital Length of Stay

based on reduced model:

```{r}
# Fit one-way ANOVA on log-transformed outcome by age_group
anova_age <- aov(log_los ~ age_group, data = df_clean)

# Tukey HSD post-hoc test
TukeyHSD(anova_age)


```

### **Conclusion:**

There is **no evidence** that hospital length of stay differs significantly by age group based on Tukey’s multiple comparisons. All group differences were small and not statistically meaningful.

We run **separate ANOVAs within levels** of `cardiohx` or `age_group` and then do Tukey on `age_group` inside each subgroup.

### Tukey HSD Post-Hoc Analysis: Age Group Differences in Length of Stay Among Patients With Cardiac History

based on reduced model:

```{r}
library(emmeans)

# Post-hoc comparisons of age_group within each level of cardiac history
emmeans(lm_reduced, pairwise ~ age_group | cardiohx, adjust = "tukey")


```

### **Post-Hoc Comparison Conclusion (Based on Reduced Model)**

Post-hoc comparisons of age group within each level of cardiac history were conducted using estimated marginal means with Tukey adjustment. In both subgroups—patients with and without a history of cardiovascular disease—no statistically significant pairwise differences were observed between age groups at the 0.05 significance level.

Among patients without cardiac history, mean log-transformed length of stay values ranged from 2.48 to 2.62 across age groups, with all pairwise comparisons yielding adjusted p-values above 0.30. Similarly, in patients with cardiac history, estimated means ranged from 2.32 to 2.60, and none of the age group contrasts reached statistical significance. Although patients aged 65–80 and 80+ with cardiac history exhibited lower average hospitalization durations, these differences were not statistically significant after multiple comparison adjustment.

These results suggest that **age group alone did not significantly influence hospital length of stay** within either cardiac history stratum, despite the overall interaction between age group and cardiac history being statistically significant in the main model.

### Plot of model-adjusted means and confidence intervals

```{r}
library(emmeans)
library(ggplot2)

# Generate estimated marginal means
emm <- emmeans(lm_reduced, ~ age_group * cardiohx)

# Create ggplot object
emm_plot <- plot(emm, comparisons = TRUE, by = "cardiohx")

# Add a custom title
emm_plot + ggtitle("Estimated Marginal Means:\nAge Group × Cardiac History on Log Length of Stay")


```

### **Interpretation:**

This plot displays the estimated marginal means of log-transformed hospital length of stay across age groups, stratified by cardiac history. While the differences between age groups are not statistically significant after Tukey adjustment (as shown by overlapping confidence intervals), the pattern varies by cardiac history status:

-   **For patients without cardiac history (cardiohx = 0):** Mean log length of stay is relatively stable across age groups, with minor fluctuations.

-   **For patients with cardiac history (cardiohx = 1):** There is a visible decrease in mean log length of stay from younger to older age groups, suggesting a potential trend where older patients with cardiac history are discharged sooner.

These trends visually support the **significant interaction** found in the ANOVA, indicating that the effect of age group on length of stay depends on cardiac history status—even if no specific age group pair reaches significance in post-hoc comparisons.

### Interaction Plot for Age Group × Cardiac History → Length of Stay

```{r}
library(dplyr)
library(ggplot2)

# Summarize log-transformed mean length of stay by age group and cardiac history
plot_data <- df_clean %>%
  group_by(age_group, cardiohx) %>%
  summarise(mean_log_los = mean(log_los, na.rm = TRUE), .groups = "drop")

# Plot interaction
ggplot(plot_data, aes(x = age_group, y = mean_log_los, color = cardiohx, group = cardiohx)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    title = "Interaction Plot: Age Group × Cardiac History → Log Length of Stay",
    x = "Age Group",
    y = "Mean Log Length of Stay",
    color = "Cardiac History"
  ) +
  theme_minimal()

```

### Interpretation of Interaction Plot:

The plot illustrates a **statistically significant interaction** between **age group** and **cardiac history** (*p* = 0.0388), indicating that the effect of age on hospital length of stay depends on whether the patient has a history of cardiovascular disease.

-   For patients **without cardiac history** (red line), length of stay is relatively stable but peaks in the **65–80** age group.

-   For patients **with cardiac history** (blue line), length of stay **decreases sharply with age**, especially after 50–65.

-   This crossover pattern supports a meaningful interaction: the direction and strength of the age effect **differs by cardiac history group**.

### Boxplot for Age Group × Cardiac History

```{r}
library(ggplot2)

ggplot(df_clean, aes(x = age_group, y = log_los, fill = age_group)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.5, alpha = 0.8) +
  facet_wrap(~ cardiohx, labeller = label_both) +
  labs(
    title = "Log-Transformed Length of Stay by Age Group and Cardiac History",
    x = "Age Group",
    y = "Log Length of Stay"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Log Length of Stay by Age Group and Cardiac History:**

This plot reveals that patients with a history of cardiovascular disease (right panel) generally show a **steeper decline** in length of stay with increasing age, particularly in the 65–80 group. In contrast, patients **without cardiac history** (left panel) exhibit **more stable median lengths of stay** across age groups, with slightly elevated values in the \<50 and 65–80 categories. The variability (IQR and outliers) appears wider in some younger groups, especially for those without cardiac history.

### Final Conclusion

This study utilized the Right Heart Catheterization (RHC) dataset from the SUPPORT study to investigate how age group and cardiac history influence hospital length of stay among critically ill patients. After verifying ANOVA assumptions and applying a log transformation to correct for skewness, a three-way ANOVA model was initially fitted including sex, age group, and cardiac history. However, model comparison via Type II ANOVA and information criteria (AIC and BIC) revealed that a reduced two-way ANOVA model — including only age group, cardiac history, and their interaction — offered equivalent or better model fit with greater parsimony.

Type II ANOVA was chosen instead of the traditional (Type I) approach because the study design is unbalanced, meaning that group sizes vary across the combinations of sex, age group, and cardiac history. Type II ANOVA provides more accurate tests of main effects and interactions in such settings by accounting for the unequal distribution of observations.

The reduced Type II ANOVA model identified a statistically significant interaction between age group and cardiac history (p = 0.0388), indicating that the effect of age on length of stay depends on cardiac history status. Specifically, length of stay decreased with age among patients with cardiac history, while those without cardiac history showed relatively stable or increasing patterns in older age groups. These trends were confirmed through interaction plots and supported by Tukey HSD post-hoc comparisons, although no pairwise differences were statistically significant.

To better visualize these results, a plot of model-adjusted means and confidence intervals was generated using the estimated marginal means from the reduced model. This visualization clearly displayed the pattern of log-transformed length of stay across age groups and cardiac history strata, aligning with the statistical findings and highlighting the nature of the interaction effect.

Visualizations, including boxplots and interaction plots, further illustrated this interaction. While sex did not contribute significantly to explaining variation in length of stay, the age group × cardiac history interaction emerged as a meaningful and interpretable driver of variation in hospital resource use.

In summary, this study concludes that the interaction between age group and cardiac history significantly affects hospital length of stay. A reduced two-way Type II ANOVA model provides a more efficient and statistically supported explanation than a full model including sex. These findings highlight the importance of considering both age and comorbid conditions when analyzing healthcare utilization in critical care populations.

------------------------------------------------------------------------
